The CrypTool Script: \\ Cryptography, Mathematics and More \\ \vspace{30pt} {\large Background reading\\%*[-8pt] for the free e-learning program CrypTool\\ (10th edition -- distributed with CrypTool version 1.4.30)} \vspace{60pt}

### 2016-02-15 script/en/chapters/numbertheory.tex
Corrected typos and overfull hbox; congruent index (no redundancy in wording; adjectives in lowercase in German; all normal words in lowercase in English)

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\section{Mathematics and cryptography}
A large proportion of modern, asymmetric cryptography\index{Encryption!asymmetric} is based on mathematical knowledge -- on the properties 
(``laws'') of whole numbers, which are investigated in elementary \index{Number theory!elementary} number 
theory. Here, the word ``elementary'' means that questions raised in number theory are essentially rooted in the set 
of natural and whole numbers.

Further mathematical disciplines currently used in cryptography include
(see \cite[p. 2]{nt:Bauer1995}, \cite[p. 3]{nt:Bauer2000}) :
\begin{itemize}
    \item Group theory\index{Group}
    \item Combination theory
    \item Complexity theory
    \item Stochastic (ergodic theory)
    \item Information theory.
\end{itemize}

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\section{Introduction to number theory} \index{Number theory!introduction}

Number theory arose from interest in positive whole numbers $1, 2, 3, 4, \cdots$, also referred to as the set of natural numbers 
\index{Number!natural} {\em natural numbers} $\mathbb{N}$. These are the first mathematical constructs 
used by human civilization. According to Kronecker\footnote{Leopold
  Kronecker, German mathematician, Dec 7, 1823 $-$ Dec 29, 1891}\index{Kronecker, Leopold}, they are a creation of God. 
In Dedekind's\footnote{Julius Wilhelm Richard Dedekind,
  German mathematician, Oct 6, 1831 $-$ Feb 12, 1916.}\index{Dedekind, Julius} opinion, they are a creation of the human intellect. Dependent upon one's ideology,
this is an unsolvable contradiction or one and the same thing.

@@

Up until the mid 20th century, number theory was considered to be the purest area of mathematics, an area that 
had no practical use in the real world. This changed with the development of computers and digital 
communication, as number theory was able to provide several unexpected solutions to real-life tasks. At the same 
time, advances in information technology allowed specialists in number theory to make huge progress in 
factorizing large numbers, finding new prime numbers, testing (old) conjectures and solving numerical problems 
that were previously impossible to solve. Modern number theory \index{Number theory!modern} is made up of 
areas such as:
\begin{itemize}
    \item Elementary number theory
    \item Algebraic number theory
    \item Analytic number theory
    \item Geometric number theory
    \item Combinatorial number theory
    \item Numeric number theory
    \item Probability theory.
\end{itemize}

@@

% --------------------------------------------------------------------------
\vskip +40 pt
\subsection{Convention}
\index{Convention}
Unless stated otherwise: 
\begin{itemize}
\item The letters $a, b, c, d, e, k, n, m, p, q$ are used to present whole numbers.
\item The letters $i$ ~\mbox{and} $j$ represent natural numbers. 
\item The letters $p$ always represents a prime number.
\item The sets $\mathbb{N} = \{ 1, 2, 3, \cdots \}$ and $\mathbb{Z} =\{ \cdots, -3, -2, -1, 0, 1, 2, 3, \cdots \}$ 
are the {\em natural} and {\em whole} numbers respectively.
\end{itemize}

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\section{Prime numbers and the first fundamental theorem of elementary number theory}
\index{Number theory!elementary}
Many of the problems in elementary number theory are concerned with prime numbers (see chapter~\ref{Label_Kapitel_Primes}).

@@

\begin{definition}\label{def-zth-prime} \index{Prime number}
{\bf Prime numbers} are natural numbers greater than 1 that can only be divided by 1 and themselves.
\end{definition}

@@

\begin{definition}\label{def-zth-composite} \index{Number!composite}
Natural numbers greater than $1$ that are not prime are called {\bf composite numbers}. These have at least two 
factors other than $1$.
\end{definition}

Examples of the decomposition of such numbers into prime 
factors\index{Prime factor!decomposition}:
\begin{eqnarray*}
4 & = & 2*2  \nonumber \\
6 & = & 2*3  \nonumber \\
91 & = & 7*13  \nonumber \\
161 & = & 7*23  \nonumber \\
767 & = & 13*59  \nonumber \\
1029 & = & 3 * 7^3  \nonumber \\
5324 & = & 22 * 11^3.  \nonumber 
\end{eqnarray*}

@@

This is the claim of the 1st {\em fundamental theorem of number theory} 
(= fundamental theorem of arithmetic = fundamental building block of all positive integers). This was 
formulated precisely for the first time by Carl Friedrich Gauss in his Disquisitiones Arithmeticae (1801).  
\index{Number theory!fundamental theorem}  \index{Gauss, Carl Friedrich}

@@

For example, $60 = 2*2*3*5 = 2^2*3*5$. And this --- other than changing the order of the factors --- is the only 
way in which the number $60$ can be factorized\index{Prime factor}.

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\newpage
\section[Divisibility, modulus and remainder classes]
           {Divisibility, modulus and remainder classes\footnotemark}
\footnotetext{%
    \index{NT, Learning Tool for Number Theory}%
    \index{Educational tool NT}%
    With the educational tool for number theory {\bf NT} you can have a
    playful view at the calculation with congruences, discussed in this and
    the next chapter (see NT learning unit 2.1, pages 2-9/40).\\
    NT can be called in CrypTool\index{CrypTool} via the menu path
    {\bf Indiv. Procedures \textbackslash{} Number Theory Interactive
    \textbackslash{} Learning Tool for Number Theory}.
    See appendix~\ref{s:appendix-Learn-NT}.\\
    CT2\index{CrypTool 2} contains a visualisation of these methods within
    the tutorial ``{\bf World of Primes}''.
}
\index{Modulus} \index{Divisibility}

@@

\begin{definition}\label{def-zth-divisibility} \index{Divisibility}
A whole number $n$ is {\bf divisible} by a whole number $d$ if the
quotient $n/d$ is a whole number $c$ such that $n = c * d$.
\end{definition}

$n$ is called a {\em multiple} of $d$, whereas $d$ is called a \index{Divisor}
{\em divisor} or \index{Factor} {\em factor} of $n$.

@@

% --------------------------------------------------------------------------
\subsection{The modulo operation -- working with congruence} \index{Congruence}

@@

\begin{definition}\label{def-zth-remainder} \index{Remainder class}
The {\bf remainder class r modulo m} is the set of all whole numbers a that
have the same remainder $r$ when divided by $m$.
\end{definition}

@@

\begin{definition}\label{def-zth-congruence} \index{Congruence}
Two numbers $a, b \in \mathbb{N}$  are said to be congruent modulo $m \in 
\mathbb{N}$ if and only if they have the same remainder when divided by $m$.
\end{definition}

@@

From algebra theorems it follows that essential parts of the conventional
calculation rules are kept when we proceed to modular calculations over
a basic set $\mathbb{Z}$.  For example, addition remains commutative. 
The same goes for multiplication modulo $m$. The result of a
division\footnote{%
The division modulo $m$\index{Division modulo $n$} is only defined for
numbers co-prime\index{Number!co-prime} to $m$ because other numbers
have the same property as zero (this means there is no inverse number).
See law number 6 {\bf existence of an inverse element}.
See footnote \ref{ftn-mod6} in chapter~\ref{addmult} and
table~\ref{mulmod6} in chapter~\ref{add-and-mult-inverses}.
\label{ftn-res-divmodn}\label{ftn-zth-divmodn}
} is not a fraction but rather a whole number between $0$ and $m-1$.

The known laws apply:
\begin{itemize}
\item[\bf 1.] {\bf Associative law:}\index{Associative law} \\ 
    $((a+b) + c) {\rm ~(mod~ } m) \equiv  (a + (b+c)) {\rm ~(mod~ } m).$ \\
    $((a*b) * c) {\rm ~(mod~ } m) \equiv  (a * (b*c)) {\rm ~(mod~ } m).$
\item[\bf 2.] {\bf Commutative law:} \index{Commutative law}\\
    $(a+b) {\rm ~(mod~ } m) \equiv  (b+a) {\rm ~(mod~ } m).$ \\
     $(a*b) {\rm ~(mod~ } m) \equiv  (b*a) {\rm ~(mod~ } m).$
\end{itemize}
The associative law and the commutative law apply to both addition and multiplication.
\begin{itemize}
\item[\bf 3.] {\bf Distributive law:} \index{Distributive law}\\
    $ (a * (b+c)) {\rm ~(mod~ } m) \equiv  (a*b + a*c) {\rm ~(mod~ } m).$ 
\item[\bf 4.] {\bf Reducibility:} \index{Reducibility} \\
    $(a+b) {\rm ~(mod~} m) \equiv  (a {\rm ~(mod~ } m) + b {\rm ~(mod~ } m)) {\rm ~(mod~} m).$ \\  
    $(a*b) {\rm ~(mod~} m) \equiv  (a {\rm ~(mod~ } m) * b {\rm ~(mod~ } m)) {\rm ~(mod~} m).$ \\
    When adding or multiplying the order in which the modulo operation is performed does not matter.
\end{itemize}

\begin{itemize}
\item[\bf 5.] {\bf Existence of an identity (neutral element):} \index{Identity}\\
    $(a + 0) {\rm ~(mod~ } m) \equiv  (0 + a) {\rm ~(mod~ } m) \equiv  a {\rm ~(mod~ } m).$  \\
    $(a * 1) {\rm ~(mod~ } m) \equiv  (1 * a) {\rm ~(mod~ } m) \equiv  a {\rm ~(mod~ } m).$

@@

\item {\bf Additive inverse}\index{Inverse!additive}\\
   For all whole numbers $a$ and $m$ there exists a whole number $-a$ such
   that: \\
   $(a + (-a)) {\rm ~(mod~}m) \equiv  0 {\rm ~(mod~ } m)$
   % $(a + (-a)) {\rm ~(mod~}m) \equiv  0 {\rm ~(mod~ } m)$ \quad (additive inverse)

\item {\bf Multiplicative inverse modulo a prime p}\\
    For each whole number $a$ (with $a \not\equiv 0 {\rm ~(mod~ } p$) and $p$
    prime) there exists a whole number $a^{-1}$, such that:~~~ 
    $(a * a^{-1}) {\rm ~(mod~ } p) \equiv 1 {\rm ~(mod~}p)$
    % \quad (multiplicative inverse)
    \index{Inverse!multiplicative}

@@

\item[\bf 7.] \index{Closeness} {\bf Closeness}\footnote{\label{ftn-zth-closed}The
   property of closeness is always defined in relation to an operation in a set.
   See chapter~\ref{nt:NumberTheory_Appendix_B} ``\nameref{nt:NumberTheory_Appendix_B}''.}: \\
      %%% \hyperlink{nt:NumberTheory_Appendix_B}{Appendix B of this chapter}.}:   \\
     $a, b \in G  \Longrightarrow  ( a + b ) \in G.$ \\
    $a, b \in G  \Longrightarrow  ( a * b ) \in G.$

\item[\bf 8.] \index{Transitivity} {\bf Transitivity}:

@@

In general mathematicians investigate ``Structures''\index{Structure}. They ask e.g.\ at $ a * x \equiv b $ mod $m$,
which values $x$ can take for given values of $a, ~b, ~m.$

@@

% --------------------------------------------------------------------------
\subsection{Addition and multiplication}
\label{addmult}
\index{Addition} \index{Multiplication}

@@

\item the multiplication tables\footnote{\label{ftn-mod6}%
Comment on modulo division:\index{Division modulo $n$}\\
Due to the special role of zero as the identity for addition, division by zero
is not permitted.\\
For all $a$ it is $a*0=0,$ because $a*0=a*(0+0)=a*0+a*0.$ Obviously $0$ has no
inverse regarding the multiplication, because if there would be one, it must
be $ 0= 0*0^{-1} =1.$ Also see footnote~\ref{ftn-zth-divmodn}
in chapter~\ref{Laws-modular-calcs}.}
for mod $5$ (table~\ref{mulmod5}) and mod $6$ (table~\ref{mulmod6}).

@@

% --------------------------------------------------------------------------
\subsection{Additive and multiplicative inverses}
\label{add-and-mult-inverses}\label{multmodn}
\index{Inverse!additive} \index{Inverse!multiplicative}

@@

However, this is not the case for modular multiplication (important theorem):
\begin{theorem}\label{thm-zth-multinv}
A natural number $a$ from the set $\{1, \cdots, m-1\}$ has one modular
multiplicative inverse if and only if this number and the modulo $m$ are
co-prime\footnote{Two whole numbers $a$ and $b$ are co-prime\index{Number!co-prime} if and only if ${\rm gcd}(a, b) = 1$.\\
If $p$ is prime and $a$ is a random whole number that is not a multiple of
$p$, then $p$ and a are co-prime.\\
Further name to the topic co-prime (with $a_i \in \mathbb{Z}, i=1, \cdots, n$):
\begin{enumerate}
\item $a_1,a_2, \cdots, a_n$ are {\em relatively prime}
\index{Prime number!relative prime}\index{Number!relative prime},
if $ \mbox{gcd}(a_1, \cdots , a_n) =1.$
\item An even stronger request for more than two numbers is :\\
     $a_1, \cdots , a_n$ are {\em in pairs relatively prime}, if for all $i=1, \cdots, n$ and 
$j=1, \cdots , n$ with $ i \neq j $: $ \mbox{gcd} (a_i, a_j) =1. $
\end{enumerate}
\begin{example}{:}\\
$2,3,6 $ are relatively prime, because $ \mbox{gcd} (2,3,6)=1.$
\end{example}
They are not in pairs relatively prime, because
$ \mbox{gcd} (2,6)=2>1.$}\index{gcd}, in other words if $a$ and $m$ have 
no common prime factors.
\end{theorem}
Since $m=5$ is prime, the numbers $1$ to $4$ are relatively prime to $5$ and {\bf each} of these numbers has a 
multiplicative inverse in mod $5$.

@@

Cryptographically not only the unique nature of the inverse is important, but
also that the set of possible values has been exhausted.
\begin{theorem}\label{thm-zth-exhperm}
For $a,i\in \{1, \cdots, m-1 \}$ with ${\rm~gcd} (a,m)=1),$ then the product $a*i {\rm ~mod ~} m $ takes for a 
certain number $a$ all values from $\{1, \cdots ,m-1 \} $ (exhaustive permutation \index{Permutation} of the length $m-1$).\footnote{See
also theorem~\ref{thm-zth-ordp} in \hyperlink{Chapter_ElementaryNT_9}
{chapter~\ref{MultOrdPrimitveRoot}, Multiplicative order and primitive roots}.}
\end{theorem}

@@

Between $i=1, \cdots, m$, all values between $0, \cdots, m-1$ occur for the
remainders, because both $5$ and $6$ are also relatively prime
\index{Prime number!relative prime}\index{Number!relative prime}
to the modulus $m=17$.
\enlargethispage{0.5cm}

@@

\begin{table}[ht]
\begin{center}                                                                  
% \begin{tabular}{|l@{\:}||c|c|c|c|c|c|c|c|c|c|c||c|c|c|c|c|c|c|} % Das wird im Englischen zu breit
% \begin{tabular}{|l||@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}||@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|}
% Das bewirkt, dass alle Spalten nur minimale Breite haben.
\begin{tabular}{|l||c|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|c|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}||c|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|}
\hline 
i $\Rightarrow$     & 1  & 2  & 3  & 4  & 5  & 6  & 7  & 8  & 9 & 10 & 11 & 12 & 13 & 14 & 15 & 16  & 17 & 18 \\
\hline 
\hline 
$5*i$               & 5 & 10 & 15 & 20 & 25 & 30 & 35 & 40 & 45 & 50 & 55 & 60 & 65 & 70 & 75 & 80 & 85  & 90 \\
remainder                 & 5 & 10  & 3  & 8  & {\bf 1}  & 6 & 11  & 4  & 9  & 2  & 7  & 0  & 5 & 10  & 3 & 8   & 1   & 6 \\
\hline 
$6*i$               & 6 & 12 & 18 & 24 & 30 & 36 & 42 & 48 & 54 & 60 & 66 & 72 & 78 & 84 & 90 & 96 & 102 & 108 \\
remainder                 & 6  & 0  & 6  & 0  & 6  & 0  & 6  & 0  & 6  & 0  & 6  & 0  & 6  & 0  & 6  & 0   & 6   & 0 \\
\hline 
\end{tabular}
\end{center} 
\caption{Multiplication table modulo $12$ (for $a=5$ and $a=6$)}
\label{mulmod12}
\end{table}
We have calculated $(5 * i)$ (mod $12$)  and  $(6*i)$ (mod $12$).
Between $i=1, \cdots, m$, not all values between $0, \cdots, m-1$ occur and $6$ does not have an inverse mod $12$, 
because $6$ and the modulus $m=12$ are not co-prime\index{Number!co-prime}.

@@

% --------------------------------------------------------------------------
\subsection{Raising to the power} \index{Raising to the power} 
In modular arithmetic, raising to the power is defined as repeated
multiplication -- as usual. With small exceptions we can even apply the usual
rules, such as:
\begin{eqnarray*}
a^{b+c} & = & a^b * a^c,  \nonumber \\
(a^b)^c & = & a^{b*c} = a^{c*b} = (a^c)^b \nonumber
\end{eqnarray*}

>>>>

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\section{Mathematics and cryptography}
A large proportion of modern, asymmetric cryptography\index{encryption!asymmetric} is based on mathematical knowledge -- on the properties 
(``laws'') of whole numbers, which are investigated in elementary \index{number theory!elementary} number 
theory. Here, the word ``elementary'' means that questions raised in number theory are essentially rooted in the set 
of natural and whole numbers.

Further mathematical disciplines currently used in cryptography include
(see \cite[p. 2]{nt:Bauer1995}, \cite[p. 3]{nt:Bauer2000}) :
\begin{itemize}
    \item Group theory\index{group}
    \item Combination theory
    \item Complexity theory\index{complexity}
    \item Stochastic (ergodic theory)
    \item Information theory.
\end{itemize}

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\section{Introduction to number theory} \index{number theory!introduction}

Number theory arose from interest in positive whole numbers $1, 2, 3, 4, \cdots$, also referred to as the set of natural numbers 
\index{number!natural} {\em natural numbers} $\mathbb{N}$. These are the first mathematical constructs 
used by human civilization. According to Kronecker\footnote{Leopold
  Kronecker, German mathematician, Dec 7, 1823 $-$ Dec 29, 1891}\index{Kronecker, Leopold}, they are a creation of God. 
In Dedekind's\footnote{Julius Wilhelm Richard Dedekind,
  German mathematician, Oct 6, 1831 $-$ Feb 12, 1916.}\index{Dedekind, Julius} opinion, they are a creation of the human intellect. Dependent upon one's ideology,
this is an unsolvable contradiction or one and the same thing.

@@

Up until the mid 20th century, number theory was considered to be the purest area of mathematics, an area that 
had no practical use in the real world. This changed with the development of computers and digital 
communication, as number theory was able to provide several unexpected solutions to real-life tasks. At the same 
time, advances in information technology allowed specialists in number theory to make huge progress in 
factorizing large numbers, finding new prime numbers, testing (old) conjectures and solving numerical problems 
that were previously impossible to solve. Modern number theory \index{number theory!modern} is made up of 
areas such as:
\begin{itemize}
    \item Elementary number theory
    \item Algebraic number theory
    \item Analytic number theory
    \item Geometric number theory
    \item Combinatorial number theory
    \item Numeric number theory
    \item Probability theory.
\end{itemize}

@@

% --------------------------------------------------------------------------
\vskip +40 pt
\subsection{Convention}
\index{convention}
Unless stated otherwise: 
\begin{itemize}
\item The letters $a, b, c, d, e, k, n, m, p, q$ are used to present whole numbers.
\item The letters $i$ ~\mbox{and} $j$ represent natural numbers. 
\item The letters $p$ always represents a prime number.
\item The sets $\mathbb{N} = \{ 1, 2, 3, \cdots \}$ and $\mathbb{Z} =\{ \cdots, -3, -2, -1, 0, 1, 2, 3, \cdots \}$ 
are the {\em natural} and {\em whole} numbers respectively.
\end{itemize}

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\section{Prime numbers and the first fundamental theorem of elementary number theory}
\index{number theory!elementary}
Many of the problems in elementary number theory are concerned with prime numbers (see chapter~\ref{Label_Kapitel_Primes}).

@@

\begin{definition}\label{def-zth-prime} \index{prime number}
{\bf Prime numbers} are natural numbers greater than 1 that can only be divided by 1 and themselves.
\end{definition}

@@

\begin{definition}\label{def-zth-composite} \index{number!composite}
Natural numbers greater than $1$ that are not prime are called {\bf composite numbers}. These have at least two 
factors other than $1$.
\end{definition}

Examples of the decomposition of such numbers into prime 
factors\index{prime factor!decomposition}:
\begin{eqnarray*}
4 & = & 2*2  \nonumber \\
6 & = & 2*3  \nonumber \\
91 & = & 7*13  \nonumber \\
161 & = & 7*23  \nonumber \\
767 & = & 13*59  \nonumber \\
1029 & = & 3 * 7^3  \nonumber \\
5324 & = & 22 * 11^3.  \nonumber 
\end{eqnarray*}

@@

This is the claim of the 1st {\em fundamental theorem of number theory} 
(= fundamental theorem of arithmetic = fundamental building block of all positive integers). This was 
formulated precisely for the first time by Carl Friedrich Gauss in his Disquisitiones Arithmeticae (1801).  
\index{number theory!fundamental theorem}  \index{Gauss, Carl Friedrich}

@@

For example, $60 = 2*2*3*5 = 2^2*3*5$. And this --- other than changing the order of the factors --- is the only 
way in which the number $60$ can be factorized\index{prime factor}.

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\newpage
\section[Divisibility, modulus and remainder classes]
           {Divisibility, modulus and remainder classes\footnotemark}
\footnotetext{%
    \index{NT, Learning Tool for Number Theory}%
    \index{educational tool NT}%
    With the educational tool for number theory {\bf NT} you can have a
    playful view at the calculation with congruences, discussed in this and
    the next chapter (see NT learning unit 2.1, pages 2-9/40).\\
    NT can be called in CrypTool\index{CrypTool} via the menu path
    {\bf Indiv. Procedures \textbackslash{} Number Theory Interactive
    \textbackslash{} Learning Tool for Number Theory}.
    See appendix~\ref{s:appendix-Learn-NT}.\\
    CT2\index{CrypTool 2} contains a visualisation of these methods within
    the tutorial ``{\bf World of Primes}''.
}
\index{modulus} \index{divisibility}

@@

\begin{definition}\label{def-zth-divisibility} \index{divisibility}
A whole number $n$ is {\bf divisible} by a whole number $d$ if the
quotient $n/d$ is a whole number $c$ such that $n = c * d$.
\end{definition}

$n$ is called a {\em multiple} of $d$, whereas $d$ is called a \index{divisor}
{\em divisor} or \index{factor} {\em factor} of $n$.

@@

% --------------------------------------------------------------------------
\subsection{The modulo operation -- working with congruence} \index{congruence}

@@

\begin{definition}\label{def-zth-remainder} \index{remainder class}
The {\bf remainder class r modulo m} is the set of all whole numbers a that
have the same remainder $r$ when divided by $m$.
\end{definition}

@@

\begin{definition}\label{def-zth-congruence} \index{congruence}
Two numbers $a, b \in \mathbb{N}$  are said to be congruent modulo $m \in 
\mathbb{N}$ if and only if they have the same remainder when divided by $m$.
\end{definition}

@@

From algebra theorems it follows that essential parts of the conventional
calculation rules are kept when we proceed to modular calculations over
a basic set $\mathbb{Z}$.  For example, addition remains commutative. 
The same goes for multiplication modulo $m$. The result of a
division\footnote{%
The division modulo $m$\index{division modulo $n$} is only defined for
numbers co-prime\index{number!co-prime} to $m$ because other numbers
have the same property as zero (this means there is no inverse number).
See law number 6 {\bf existence of an inverse element}.
See footnote \ref{ftn-mod6} in chapter~\ref{addmult} and
table~\ref{mulmod6} in chapter~\ref{add-and-mult-inverses}.
\label{ftn-res-divmodn}\label{ftn-zth-divmodn}
} is not a fraction but rather a whole number between $0$ and $m-1$.

The known laws apply:
\begin{itemize}
\item[\bf 1.] {\bf Associative law:}\index{associative law} \\ 
    $((a+b) + c) {\rm ~(mod~ } m) \equiv  (a + (b+c)) {\rm ~(mod~ } m).$ \\
    $((a*b) * c) {\rm ~(mod~ } m) \equiv  (a * (b*c)) {\rm ~(mod~ } m).$
\item[\bf 2.] {\bf Commutative law:} \index{commutative law}\\
    $(a+b) {\rm ~(mod~ } m) \equiv  (b+a) {\rm ~(mod~ } m).$ \\
     $(a*b) {\rm ~(mod~ } m) \equiv  (b*a) {\rm ~(mod~ } m).$
\end{itemize}
The associative law and the commutative law apply to both addition and multiplication.
\begin{itemize}
\item[\bf 3.] {\bf Distributive law:} \index{distributive law}\\
    $ (a * (b+c)) {\rm ~(mod~ } m) \equiv  (a*b + a*c) {\rm ~(mod~ } m).$ 
\item[\bf 4.] {\bf Reducibility:} \index{reducibility} \\
    $(a+b) {\rm ~(mod~} m) \equiv  (a {\rm ~(mod~ } m) + b {\rm ~(mod~ } m)) {\rm ~(mod~} m).$ \\  
    $(a*b) {\rm ~(mod~} m) \equiv  (a {\rm ~(mod~ } m) * b {\rm ~(mod~ } m)) {\rm ~(mod~} m).$ \\
    When adding or multiplying the order in which the modulo operation is performed does not matter.
\end{itemize}

\begin{itemize}
\item[\bf 5.] {\bf Existence of an identity (neutral element):} \index{identity}\\
    $(a + 0) {\rm ~(mod~ } m) \equiv  (0 + a) {\rm ~(mod~ } m) \equiv  a {\rm ~(mod~ } m).$  \\
    $(a * 1) {\rm ~(mod~ } m) \equiv  (1 * a) {\rm ~(mod~ } m) \equiv  a {\rm ~(mod~ } m).$

@@

\item {\bf Additive inverse}\index{inverse!additive}\\
   For all whole numbers $a$ and $m$ there exists a whole number $-a$ such
   that: \\
   $(a + (-a)) {\rm ~(mod~}m) \equiv  0 {\rm ~(mod~ } m)$
   % $(a + (-a)) {\rm ~(mod~}m) \equiv  0 {\rm ~(mod~ } m)$ \quad (additive inverse)

\item {\bf Multiplicative inverse modulo a prime p}\\
    For each whole number $a$ (with $a \not\equiv 0 {\rm ~(mod~ } p$) and $p$
    prime) there exists a whole number $a^{-1}$, such that:~~~ 
    $(a * a^{-1}) {\rm ~(mod~ } p) \equiv 1 {\rm ~(mod~}p)$
    % \quad (multiplicative inverse)
    \index{inverse!multiplicative}

@@

\item[\bf 7.] \index{closeness} {\bf Closeness}\footnote{\label{ftn-zth-closed}The
   property of closeness is always defined in relation to an operation in a set.
   See chapter~\ref{nt:NumberTheory_Appendix_B} ``\nameref{nt:NumberTheory_Appendix_B}''.}: \\
      %%% \hyperlink{nt:NumberTheory_Appendix_B}{Appendix B of this chapter}.}:   \\
     $a, b \in G  \Longrightarrow  ( a + b ) \in G.$ \\
    $a, b \in G  \Longrightarrow  ( a * b ) \in G.$

\item[\bf 8.] \index{transitivity} {\bf Transitivity}:

@@

In general mathematicians investigate ``Structures''\index{structure}. They ask e.g.\ at $ a * x \equiv b $ mod $m$,
which values $x$ can take for given values of $a, ~b, ~m.$

@@

% --------------------------------------------------------------------------
\subsection{Addition and multiplication}
\label{addmult}
\index{addition} \index{multiplication}

@@

\item the multiplication tables\footnote{\label{ftn-mod6}%
Comment on modulo division:\index{division modulo $n$}\\
Due to the special role of zero as the identity for addition, division by zero
is not permitted.\\
For all $a$ it is $a*0=0,$ because $a*0=a*(0+0)=a*0+a*0.$ Obviously $0$ has no
inverse regarding the multiplication, because if there would be one, it must
be $ 0= 0*0^{-1} =1.$ Also see footnote~\ref{ftn-zth-divmodn}
in chapter~\ref{Laws-modular-calcs}.}
for mod $5$ (table~\ref{mulmod5}) and mod $6$ (table~\ref{mulmod6}).

@@

% --------------------------------------------------------------------------
\subsection{Additive and multiplicative inverses}
\label{add-and-mult-inverses}\label{multmodn}
\index{inverse!additive} \index{inverse!multiplicative}

@@

However, this is not the case for modular multiplication (important theorem):
\begin{theorem}\label{thm-zth-multinv}
A natural number $a$ from the set $\{1, \cdots, m-1\}$ has one modular
multiplicative inverse if and only if this number and the modulo $m$ are
co-prime\footnote{Two whole numbers $a$ and $b$ are co-prime\index{number!co-prime} if and only if ${\rm gcd}(a, b) = 1$.\\
If $p$ is prime and $a$ is a random whole number that is not a multiple of
$p$, then $p$ and a are co-prime.\\
Further name to the topic co-prime (with $a_i \in \mathbb{Z}, i=1, \cdots, n$):
\begin{enumerate}
\item $a_1,a_2, \cdots, a_n$ are {\em relatively prime}
\index{prime number!relative prime}\index{number!relative prime},
if $ \mbox{gcd}(a_1, \cdots , a_n) =1.$
\item An even stronger request for more than two numbers is :\\
     $a_1, \cdots , a_n$ are {\em in pairs relatively prime}, if for all $i=1, \cdots, n$ and 
$j=1, \cdots , n$ with $ i \neq j $: $ \mbox{gcd} (a_i, a_j) =1. $
\end{enumerate}
\begin{example}{:}\\
$2,3,6 $ are relatively prime, because $ \mbox{gcd} (2,3,6)=1.$
\end{example}
They are not in pairs relatively prime, because
$ \mbox{gcd} (2,6)=2>1.$}\index{gcd}, in other words if $a$ and $m$ have 
no common prime factors.
\end{theorem}
Since $m=5$ is prime, the numbers $1$ to $4$ are relatively prime to $5$ and {\bf each} of these numbers has a 
multiplicative inverse in mod $5$.

@@

Cryptographically not only the unique nature of the inverse is important, but
also that the set of possible values has been exhausted.
\begin{theorem}\label{thm-zth-exhperm}
For $a,i\in \{1, \cdots, m-1 \}$ with ${\rm~gcd} (a,m)=1),$ then the product $a*i {\rm ~mod ~} m $ takes for a 
certain number $a$ all values from $\{1, \cdots ,m-1 \} $ (exhaustive permutation \index{permutation} of the length $m-1$).\footnote{See
also theorem~\ref{thm-zth-ordp} in \hyperlink{Chapter_ElementaryNT_9}
{chapter~\ref{MultOrdPrimitveRoot}, Multiplicative order and primitive roots}.}
\end{theorem}

@@

Between $i=1, \cdots, m$, all values between $0, \cdots, m-1$ occur for the
remainders, because both $5$ and $6$ are also relatively prime
\index{prime number!relative prime}\index{number!relative prime}
to the modulus $m=17$.
\enlargethispage{0.5cm}

@@

\begin{table}[ht]
\begin{center}                                                                  
% \begin{tabular}{|l@{\:}||c|c|c|c|c|c|c|c|c|c|c||c|c|c|c|c|c|c|} % Das wird im Englischen zu breit
% \begin{tabular}{|l||@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}||@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|}
% Das bewirkt, dass alle Spalten nur minimale Breite haben.
\begin{tabular}{|l||c|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|c|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}||c|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|}
\hline 
i $\Rightarrow$     & 1  & 2  & 3  & 4  & 5  & 6  & 7  & 8  & 9 & 10 & 11 & 12 & 13 & 14 & 15 & 16  & 17 & 18 \\
\hline 
\hline 
$5*i$               & 5 & 10 & 15 & 20 & 25 & 30 & 35 & 40 & 45 & 50 & 55 & 60 & 65 & 70 & 75 & 80 & 85  & 90 \\
remainder                 & 5 & 10  & 3  & 8  & {\bf 1}  & 6 & 11  & 4  & 9  & 2  & 7  & 0  & 5 & 10  & 3 & 8   & 1   & 6 \\
\hline 
$6*i$               & 6 & 12 & 18 & 24 & 30 & 36 & 42 & 48 & 54 & 60 & 66 & 72 & 78 & 84 & 90 & 96 & 102 & 108 \\
remainder                 & 6  & 0  & 6  & 0  & 6  & 0  & 6  & 0  & 6  & 0  & 6  & 0  & 6  & 0  & 6  & 0   & 6   & 0 \\
\hline 
\end{tabular}
\end{center} 
\caption{Multiplication table modulo $12$ (for $a=5$ and $a=6$)}
\label{mulmod12}
\end{table}
We have calculated $(5 * i)$ (mod $12$)  and  $(6*i)$ (mod $12$).
Between $i=1, \cdots, m$, not all values between $0, \cdots, m-1$ occur and $6$ does not have an inverse mod $12$, 
because $6$ and the modulus $m=12$ are not co-prime\index{number!co-prime}.

@@

% --------------------------------------------------------------------------
\subsection{Raising to the power} \index{raising to the power}\index{power}
In modular arithmetic, raising to the power is defined as repeated
multiplication -- as usual. With small exceptions we can even apply the usual
rules, such as:
\begin{eqnarray*}
a^{b+c} & = & a^b * a^c,  \nonumber \\
(a^b)^c & = & a^{b*c} = a^{c*b} = (a^c)^b \nonumber
\end{eqnarray*}
